Appendix 4: species distribution model comparison between INLA and sdmTMB

### 2021-12-22 doc/appendix-binomial-inla.Rmd
Edit inla appendix

```{r packages, message=FALSE, warning=FALSE, cache=FALSE}
library(ggplot2)
library(dplyr)
library(sdmTMB)
library(INLA)
# library(inlabru)
theme_set(ggsidekick::theme_sleek())
```

In this example, we fit a presence-absence species distribution model with fixed effect intercepts for year and spatiotemporal random fields using AR1. 
This will be fit with a binomial distribution and a logit link. 
Once again we will use a data set built into the package: trawl survey data for Pacific Cod in Queen Charlotte Sound.
INLA requires sequentially numbered time steps, we start by selecting the years we will include in this model and creating the year index variable `i_year`.

@@

In this example, we will use identical priors for the range and random field SD (aka. sigma) for models built in INLA and with sdmTMB.
While being described different in each package, the numbers amount to the probability the range is below a lower limit (e.g., `P(range < 5) = 0.05`) and that the sigma is above an upper limit (e.g., `P(sigma > 5) = 0.05`).

@@

In INLA a PC-prior can be used for the temporal autoregressive parameter, i.e., the which defines the probability the autocorrelation parameter is greater than a given value (e.g., `P(cor > 0) = 0.9`), but for now sdmTMB only takes a mean and SD (e.g., `ar1_rho = normal(0, 1)`) so we will not set a prior for either model.

```{r}
h.spec <- list(rho = list(prior = 'pccor1', param = c(0, 0.9)))
```

We will start by building an identical meshes using both sdmTMB and INLA.
Here, X and Y are coordinates in UTM zone 9.
These same functions that are used internally in the sdmTMB `make_mesh()` function.

@@

```{r inla-mesh, echo = TRUE, cache=FALSE}
loc_xy <- as.matrix(d[, c("X", "Y"), drop = FALSE])
inla_mesh <- INLA::inla.mesh.create(loc_xy, refine = TRUE, cutoff = 10)
spde <- inla.spde2.pcmatern(
  mesh = inla_mesh, # an smdTMB mesh can be inserted here with sdm_mesh$mesh
    prior.range = c(range_min, prior_prob),
    prior.sigma = c(sigma_max, prior_prob)) 
```

```{r plot-inla-mesh, fig.asp=0.9, out.height="2.5in"}
plot(sdm_mesh);title(main = "sdmTMB mesh");plot(spde$mesh)
```

@@

Data preparation for INLA models includes a formulae, as well as objects made using several different functions including: `inla.spde.make.index()`, `inla.spde.make.A()`, and `inla.stack()`.
TODO: explain further, including `control.inla = list(int.strategy = "eb", strategy = "gaussian")` 
TODO: We use very weak priors on the fixed effects (`mean = 0, prec = 1/(100*100)`)?

```{r inla}
tictoc::tic("INLA")

nyear <- length(unique(d$year))

iset <- inla.spde.make.index('i', n.spde = spde$n.spde, n.group = nyear)

A <- inla.spde.make.A(mesh = inla_mesh, loc = loc_xy, group = d$i_year)

sdat <- inla.stack(
  data = list(y = d$present),
  A = list(A, 1),
  effects = list(iset, year_factor = as.factor(d$year)),
  tag = 'stdata')

formulae <- y ~ 0 + year_factor + f(i, model = spde, group = i.group,
  control.group = list(model = 'ar1'))
  # control.group = list(model = 'ar1', hyper = h.spec))

m_inla <- inla(formulae,
  data = inla.stack.data(sdat),
  control.predictor = list(compute = TRUE, A = inla.stack.A(sdat)),
  family = "binomial",
  control.family = list(link = "logit"),
  control.fixed = list(expand.factor.strategy = 'inla', mean = 0, prec = 1/(100*100)),
  verbose = FALSE,
  control.inla = list(int.strategy = "eb", strategy = "gaussian"),
  control.compute = list(config = TRUE)
  )

tictoc::toc()

summary(m_inla)
# saves fixed effect coefs
inla_coefs <- m_inla$summary.fixed[, c("mean", "sd")]
```

@@

We then fit a very similar model with `sdmTMB()`. 
We again use very weak priors on the fixed effects.

```{r sdmtmb, warning=FALSE, message=FALSE}
tictoc::tic("sdmTMB")

m <- sdmTMB(present ~ 0 + as.factor(year), 
  data = d, 
  mesh = sdm_mesh, 
  time = "year",
  spatial = "off", 
  spatiotemporal = "AR1", 
  family = binomial(link = "logit"),
  priors = sdmTMBpriors(
    matern_st = pc_matern(
      range_gt = range_min, range_prob = prior_prob, #P(range > 5) = 1-0.05
      sigma_lt = sigma_max, sigma_prob = prior_prob #P(sigma < 5) = 1-0.05
    ),
    b = normal(0, 100)
  ))

tictoc::toc()
summary(m)
tidy(m, "ran_pars", conf.int = TRUE)
# saves fixed effect coefs
sdmtmb_coefs <- tidy(m, effects = "fixed")
```

@@

With sdmTMB, we can extract parameter simulations from the joint precision matrix using `spread_sims()` or `gather_sims()` functions.

@@

post_sdmTMB2 <- post_sdmTMB %>% mutate(variable = factor(`.variable`, 
                  levels = unique(post_sdmTMB$`.variable`), 
                  labels = c( "2007", "2009", "2011", "2013", "2015", "2017", 
                                "Range", "AR1 rho", "Random Field SD")))
```

We then compare the distribution of these sdmTMB simulations to the marginal posterior densities from INLA. 
From INLA, we can extract the marginal posterior densities using `inla.smarginal()` and name them to match the variable names from the sdmTMB `gather_sims()` function.
TODO: maybe say something about why not using `inla.posterior.sample(result = m_inla, n = 400L)`?

```{r inla-densities}
# post_inla <- inla.posterior.sample(result = m_inla, n = 400L)
x <- list()
x[[1]] <- inla.smarginal(m_inla$marginals.fixed$year_factor2007) %>% 
  as_tibble() %>% mutate(variable = "as.factor.year.2007") 
x[[2]] <- inla.smarginal(m_inla$marginals.fixed$year_factor2009) %>% 
  as_tibble() %>% mutate(variable = "as.factor.year.2009") 
x[[3]] <- inla.smarginal(m_inla$marginals.fixed$year_factor2011) %>% 
  as_tibble() %>% mutate(variable = "as.factor.year.2011") 
x[[4]] <- inla.smarginal(m_inla$marginals.fixed$year_factor2013) %>% 
  as_tibble() %>% mutate(variable = "as.factor.year.2013") 
x[[5]] <- inla.smarginal(m_inla$marginals.fixed$year_factor2015) %>% 
  as_tibble() %>% mutate(variable = "as.factor.year.2015") 
x[[6]] <- inla.smarginal(m_inla$marginals.fixed$year_factor2017) %>% 
  as_tibble() %>% mutate(variable = "as.factor.year.2017") 
x[[7]] <- inla.smarginal(m_inla$marginals.hyperpar$`Range for i`) %>% 
  as_tibble() %>% mutate(variable = "range") 
x[[8]] <- inla.smarginal(m_inla$marginals.hyperpar$`Stdev for i`) %>% 
  as_tibble() %>% mutate(variable = "sigma_E") 
x[[9]] <- inla.smarginal(m_inla$marginals.hyperpar$`GroupRho for i`) %>% 
  as_tibble() %>% mutate(variable = "ar1_rho") 
post_inla1 <- do.call(rbind, x)

post_inla2 <- post_inla1 %>% mutate(variable = factor(variable, 
                  levels = unique(post_inla1$variable), 
                  labels = c( "2007", "2009", "2011", "2013", "2015", "2017", 
                                "Range", "Random Field SD", "AR1 rho")))
```

```{r compare-plot, out.width="6in"}
ggplot(post_inla2) + 
  geom_histogram(data=post_sdmTMB2, aes(.value, after_stat(density)), bins = 30, 
                 fill = "blue", alpha = 0.2) + 
  geom_line(data=post_inla2, aes(x, y), inherit.aes = F) + 
  facet_wrap(~variable, scales = "free") 
```

@@

We can predict on a the grid (`qcs_grid`) provided with the `pcod` data in the sdmTMB package to map occurrence probabilities in each year.

@@

```{r inla-pred, out.width="6in"}
idx <- inla.stack.index(sdat, 'stdata')$data
d$inla_pred <- m_inla$summary.fitted.values[idx, "mean"]
ggplot(d) + geom_point(aes(X, Y, colour = inla_pred), size = 2) + 
  facet_wrap(~year) + scale_colour_viridis_c() + coord_fixed() 
```

>>>>

```{r packages, message=FALSE, warning=FALSE, cache=FALSE}
library(ggplot2)
library(dplyr)
library(sdmTMB)
library(INLA)
theme_set(theme_light())
```
<!-- # library(inlabru) -->

In this example, we fit a presence-absence species distribution model with fixed-effect intercepts for year and spatiotemporal random fields structured as AR(1). 
We will use a binomial distribution and a logit link. 
We will use a data set built into the package: trawl survey data for Pacific cod in Queen Charlotte Sound as described in earlier appendices.
INLA requires sequentially numbered time steps, we start by selecting the years we will include in this model and creating the year index variable `i_year`:

@@

In this example, we will use Penalized Complexity (PC) priors on the range and random field standard deviation ("sigma") for models built in INLA and sdmTMB.
The values amount to the probability the range is below a lower limit (e.g., `P(range < 5) = 0.05`) and that the sigma is above an upper limit (e.g., `P(sigma > 5) = 0.05`).

@@

In INLA a PC-prior can be used for the temporal autoregressive parameter, i.e., a prior that defines the probability the autocorrelation parameter is greater than a given value (e.g., `P(cor > 0) = 0.9`), but for now, sdmTMB only takes a Gaussian prior (e.g., `ar1_rho = normal(0, 1)`).

```{r}
h.spec <- list(rho = list(prior = "pccor1", param = c(0, 0.9)))
```

We will start by building identical meshes using both sdmTMB and INLA.
Here, X and Y are coordinates in UTM zone 9.
These same functions are used internally in the sdmTMB `make_mesh()` function.

@@

```{r inla-mesh, echo = TRUE, cache=FALSE}
loc_xy <- as.matrix(d[, c("X", "Y"), drop = FALSE])
inla_mesh <- INLA::inla.mesh.create(loc_xy, refine = TRUE, cutoff = 10)
spde <- inla.spde2.pcmatern(
  mesh = inla_mesh, # an smdTMB mesh can be inserted here with sdm_mesh$mesh
  prior.range = c(range_min, prior_prob),
  prior.sigma = c(sigma_max, prior_prob)
)
```

```{r plot-inla-mesh, fig.asp=0.9, out.height="2.5in"}
plot(sdm_mesh)
title(main = "sdmTMB mesh")
plot(spde$mesh)
```

@@

Data preparation for INLA models includes a formula, as well as objects made using several different functions including: `inla.spde.make.index()`, `inla.spde.make.A()`, and `inla.stack()`.
We use `control.inla = list(int.strategy = "eb", strategy = "gaussian")` for comparability with TMB (REF).
We use weak priors on the fixed effects (`mean = 0, prec = 1/(100*100)`), which amounts to a `Normal(0, 100^2)` prior.

```{r inla}
tictoc::tic("INLA") # start timing

nyear <- length(unique(d$year))
iset <- inla.spde.make.index("i", n.spde = spde$n.spde, n.group = nyear)
A <- inla.spde.make.A(mesh = inla_mesh, loc = loc_xy, group = d$i_year)
sdat <- inla.stack(
  data = list(y = d$present),
  A = list(A, 1),
  effects = list(iset, year_factor = as.factor(d$year)),
  tag = "stdata"
)

formula <- y ~ 0 + year_factor + f(i,
  model = spde, group = i.group,
  control.group = list(model = "ar1", hyper = h.spec)
)

m_inla <- inla(formula,
  data = inla.stack.data(sdat),
  control.predictor = list(compute = TRUE, A = inla.stack.A(sdat)),
  family = "binomial",
  control.family = list(link = "logit"),
  control.fixed = list(expand.factor.strategy = "inla", mean = 0, prec = 1 / (100 * 100)),
  verbose = FALSE,
  control.inla = list(int.strategy = "eb", strategy = "gaussian"),
  control.compute = list(config = TRUE)
)
tictoc::toc() # stop timing

summary(m_inla)
inla_coefs <- m_inla$summary.fixed[, c("mean", "sd")]
```

@@

We will now fit a nearly identical model with `sdmTMB()` (the AR(1) correlation parameter prior is different).
We again use very weak priors on the fixed effects.

```{r sdmtmb, warning=FALSE, message=FALSE}
tictoc::tic("sdmTMB")
m <- sdmTMB(present ~ 0 + as.factor(year),
  data = d,
  mesh = sdm_mesh,
  time = "year",
  spatial = "off",
  spatiotemporal = "ar1",
  family = binomial(link = "logit"),
  priors = sdmTMBpriors(
    matern_st = pc_matern(
      range_gt = range_min, range_prob = prior_prob,
      sigma_lt = sigma_max, sigma_prob = prior_prob
    ),
    b = normal(0, 100),
    ar1_rho = normal(0, 2)
  )
)
tictoc::toc()
summary(m)
tidy(m, "ran_pars", conf.int = TRUE)
sdmtmb_coefs <- tidy(m, effects = "fixed")
```

@@

With sdmTMB, we can extract parameter simulations from the joint precision matrix using the `spread_sims()` or `gather_sims()` functions.

@@

post_sdmTMB2 <- post_sdmTMB %>% mutate(variable = factor(`.variable`,
  levels = unique(post_sdmTMB$`.variable`),
  labels = c(
    "2007", "2009", "2011", "2013", "2015", "2017",
    "Range", "AR1 rho", "Random Field SD"
  )
))
```

We can then compare the distribution of these sdmTMB simulations to the marginal posterior densities from INLA. 
From INLA, we can extract the marginal posterior densities using `inla.smarginal()` and name them to match the variable names from the sdmTMB `gather_sims()` function.
Alternatively, we could use `inla.posterior.sample(result = m_inla, n = 1000)` to sample from the posterior.
<!-- # post_inla <- inla.posterior.sample(result = m_inla, n = 400L) -->

```{r inla-densities}
x <- list()
x[[1]] <- inla.smarginal(m_inla$marginals.fixed$year_factor2007) %>%
  as_tibble() %>%
  mutate(variable = "as.factor.year.2007")
x[[2]] <- inla.smarginal(m_inla$marginals.fixed$year_factor2009) %>%
  as_tibble() %>%
  mutate(variable = "as.factor.year.2009")
x[[3]] <- inla.smarginal(m_inla$marginals.fixed$year_factor2011) %>%
  as_tibble() %>%
  mutate(variable = "as.factor.year.2011")
x[[4]] <- inla.smarginal(m_inla$marginals.fixed$year_factor2013) %>%
  as_tibble() %>%
  mutate(variable = "as.factor.year.2013")
x[[5]] <- inla.smarginal(m_inla$marginals.fixed$year_factor2015) %>%
  as_tibble() %>%
  mutate(variable = "as.factor.year.2015")
x[[6]] <- inla.smarginal(m_inla$marginals.fixed$year_factor2017) %>%
  as_tibble() %>%
  mutate(variable = "as.factor.year.2017")
x[[7]] <- inla.smarginal(m_inla$marginals.hyperpar$`Range for i`) %>%
  as_tibble() %>%
  mutate(variable = "range")
x[[8]] <- inla.smarginal(m_inla$marginals.hyperpar$`Stdev for i`) %>%
  as_tibble() %>%
  mutate(variable = "sigma_E")
x[[9]] <- inla.smarginal(m_inla$marginals.hyperpar$`GroupRho for i`) %>%
  as_tibble() %>%
  mutate(variable = "ar1_rho")
post_inla1 <- do.call(rbind, x)
post_inla2 <- post_inla1 %>% mutate(variable = factor(variable,
  levels = unique(post_inla1$variable),
  labels = c(
    "2007", "2009", "2011", "2013", "2015", "2017",
    "Range", "Random Field SD", "AR1 rho"
  )
))
```

```{r compare-plot, out.width="6in"}
ggplot(post_inla2) +
  geom_histogram(
    data = post_sdmTMB2, aes(.value, after_stat(density)), bins = 30,
    fill = "blue", alpha = 0.2
  ) +
  geom_line(data = post_inla2, aes(x, y), inherit.aes = F) +
  facet_wrap(~variable, scales = "free")
```

@@

We can predict on a grid (`qcs_grid`) provided with the `pcod` data in the sdmTMB package to map occurrence probabilities in each year in space.

@@

```{r inla-pred, out.width="6in"}
idx <- inla.stack.index(sdat, "stdata")$data
d$inla_pred <- m_inla$summary.fitted.values[idx, "mean"]
ggplot(d) +
  geom_point(aes(X, Y, colour = inla_pred), size = 2) +
  facet_wrap(~year) +
  scale_colour_viridis_c() +
  coord_fixed()
```
